---
title: "LAB 1 Bioinformatics"
date: "`r Sys.Date()`"
author: 
  - Hugo Morvan
output:
  pdf_document:
    latex_engine: xelatex
---

```{r}
set.seed(12345)

library(ggplot2)
library(dplyr)
```

```{r 1.2}
MM = 357
NN = 158
MN = 485
pop = 1000

p = sqrt(MM/pop)
q = sqrt(NN/pop)
pq = (MN/pop)/2

#assuming p is correct:
qe = 1-p
pqe_p = p*qe 

chi_test_p = (q-qe)^2/qe + (pq-pqe_p)^2/pqe_p
print(chi_test_p)

#assuming q is correct:
pe = 1-q
pqe_q = pe*q
chi_test_q = (p-pe)^2/pe + (pq-pqe_q)^2/pqe_q
print(chi_test_q)

p_value_p <- pchisq(chi_test_p, df=2, lower.tail = FALSE)
print(p_value_p)
p_value_q <- pchisq(chi_test_q, df=2, lower.tail = FALSE)
print(p_value_q)

```


#2.2


>MK465080.1 Branchipus schaefferi isolate ST2-2 cytochrome c oxidase subunit I (COI) gene, partial cds; mitochondrial
TCTCCTAGGAGATGACCAACTTTATAACGTCATTGTTACTGCTCACGCTTTTGTAATGATTTTCTTCATA
GTTATACCAATCCTTATTGGAGGATTTGGAAATTGATTAGTCCCTTTAATACTAGGGGCTCCTGATATGG
CTTTCCCCCGACTAAATAACTTAAGCTTTTGAATACTTCCTCCCTCATTAACTCTTCTAGTGGCCAGCTC
AATGGTAGAAAGAGGGGTAGGAACAGGATGAACAGTTTATCCACCCCTATCTGCTGCTATTGCCCATGCT
GGTCCTTCTGTTGATTTAGCAATCTTTTCACTTCACCTCGCAGGGATCTCTTCAATTTTAGGAGCTGTAA
ATTTCATTACAACTGTAATTAATATACGGCCTCATTCCATAAGATTAGACCGAATACCTTTATTTGCATG
AGCGGTAGTTATCACAGCAGTTCTTCTTCTCCTTTCTCTCCCAGTATTAGCAG

From NLM
/protein_id="QDG08361.1"
                     /translation="LLGDDQLYNVIVTAHAFVMIFFMVMPILIGGFGNWLVPLMLGAP
                     DMAFPRLNNLSFWMLPPSLTLLVASSMVESGVGTGWTVYPPLSAAIAHAGPSVDLAIF
                     SLHLAGISSILGAVNFITTVINMRPHSMSLDRMPLFAWAVVITAVLLLLSLPVLA"
                     
From transeq : 
>MK465080.1_2 Branchipus schaefferi isolate ST2-2 cytochrome c oxidase subunit I (COI) gene, partial cds; mitochondrial
LLGDDQLYNVIVTAHAFVMIFFMVMPILIGGFGNWLVPLMLGAPDMAFPRLNNLSFWMLP
PSLTLLVASSMVESGVGTGWTVYPPLSAAIAHAGPSVDLAIFSLHLAGISSILGAVNFIT
TVINMRPHSMSLDRMPLFAWAVVITAVLLLLSLPVLAX
                     
                     
from backtranseq:

Download
Using codon table Caenorhabditis elegans (nematode)

>MK465080.1_2 Branchipus schaefferi isolate ST2-2 cytochrome c oxidase subunit I (COI) gene, partial cds; mitochondrial
CTTCTTGGAGATGATCAACTTTATAATGTTATTGTTACAGCTCATGCTTTCGTTATGATT
TTCTTCATGGTTATGCCAATTCTTATTGGAGGATTCGGAAATTGGCTTGTTCCACTTATG
CTTGGAGCTCCAGATATGGCTTTCCCAAGACTTAATAATCTTTCATTCTGGATGCTTCCA
CCATCACTTACACTTCTTGTTGCTTCATCAATGGTTGAATCAGGAGTTGGAACAGGATGG
ACAGTTTATCCACCACTTTCAGCTGCTATTGCTCATGCTGGACCATCAGTTGATCTTGCT
ATTTTCTCACTTCATCTTGCTGGAATTTCATCAATTCTTGGAGCTGTTAATTTCATTACA
ACAGTTATTAATATGAGACCACATTCAATGTCACTTGATAGAATGCCACTTTTCGCTTGG
GCTGTTGTTATTACAGCTGTTCTTCTTCTTCTTTCACTTCCAGTTCTTGCTNNN


Using codon table drosophilia melanogaster

>MK465080.1_2 Branchipus schaefferi isolate ST2-2 cytochrome c oxidase subunit I (COI) gene, partial cds; mitochondrial
CTGCTGGGCGATGATCAGCTGTACAACGTGATCGTGACCGCCCACGCCTTCGTGATGATC
TTCTTCATGGTGATGCCCATCCTGATCGGCGGCTTCGGCAACTGGCTGGTGCCCCTGATG
CTGGGCGCCCCCGATATGGCCTTCCCCCGCCTGAACAACCTGAGCTTCTGGATGCTGCCC
CCCAGCCTGACCCTGCTGGTGGCCAGCAGCATGGTGGAGAGCGGCGTGGGCACCGGCTGG
ACCGTGTACCCCCCCCTGAGCGCCGCCATCGCCCACGCCGGCCCCAGCGTGGATCTGGCC
ATCTTCAGCCTGCACCTGGCCGGCATCAGCAGCATCCTGGGCGCCGTGAACTTCATCACC
ACCGTGATCAACATGCGCCCCCACAGCATGAGCCTGGATCGCATGCCCCTGTTCGCCTGG
GCCGTGGTGATCACCGCCGTGCTGCTGCTGCTGAGCCTGCCCGTGCTGGCCNNN

Using codon table drosophilia melanogaster (high)

>MK465080.1_2 Branchipus schaefferi isolate ST2-2 cytochrome c oxidase subunit I (COI) gene, partial cds; mitochondrial
CTGCTGGGCGACGACCAGCTGTACAACGTGATCGTGACCGCCCACGCCTTCGTGATGATC
TTCTTCATGGTGATGCCCATCCTGATCGGCGGCTTCGGCAACTGGCTGGTGCCCCTGATG
CTGGGCGCCCCCGACATGGCCTTCCCCCGCCTGAACAACCTGTCCTTCTGGATGCTGCCC
CCCTCCCTGACCCTGCTGGTGGCCTCCTCCATGGTGGAGTCCGGCGTGGGCACCGGCTGG
ACCGTGTACCCCCCCCTGTCCGCCGCCATCGCCCACGCCGGCCCCTCCGTGGACCTGGCC
ATCTTCTCCCTGCACCTGGCCGGCATCTCCTCCATCCTGGGCGCCGTGAACTTCATCACC
ACCGTGATCAACATGCGCCCCCACTCCATGTCCCTGGACCGCATGCCCCTGTTCGCCTGG
GCCGTGGTGATCACCGCCGTGCTGCTGCTGCTGTCCCTGCCCGTGCTGGCCNNN



#2.3

Reverse (Caenorhabditis elegans codon table):

TTAGTATTGCTTTCGTACTCGACATTGTTATTGTAATATTTCAACTAGTAGAGGTTCTTC
GTATTCACCTTGTTCGGTTAAAGGCTTAGGAGGTTATTCTTAACCGTATTGGTACTTCTT
ACCTTCGTAGGTCTTACTTTCTAATAATTCAGAACCCTTTCGGTATAGACCTCGAGGTTC
GGTAGGACAAGGTTGAGGACTAAGTTGGTAACTACTTCGTTGTTCTTCACATTCACTACC
TCGTTCTAGTTGACTACCAGGTCGTACTCGTTATCGTCGACTTTCACCACCTATTTGACA
ACATTACTTTAATTGTCGAGGTTCTTAACTACTTTAAGGTCGTTCTACTTCACTCTTTTA
GGTTCGCTTTTCACCGTAAGATAGTTCACTGTAACTTACACCAGAGTATAATTATTGACA
NNNTCGTTCTTGACCTTCACTTTCTTCTTCTTCTTGTCGACATTATTGTTGTCG

Complement:

AATCATAACGAAAGCATGAGCTGTAACAATAACATTATAAAGTTGATCATCTCCAAGAAG
CATAAGTGGAACAAGCCAATTTCCGAATCCTCCAATAAGAATTGGCATAACCATGAAGAA
TGGAAGCATCCAGAATGAAAGATTATTAAGTCTTGGGAAAGCCATATCTGGAGCTCCAAG
CCATCCTGTTCCAACTCCTGATTCAACCATTGATGAAGCAACAAGAAGTGTAAGTGATGG
AGCAAGATCAACTGATGGTCCAGCATGAGCAATAGCAGCTGAAAGTGGTGGATAAACTGT
TGTAATGAAATTAACAGCTCCAAGAATTGATGAAATTCCAGCAAGATGAAGTGAGAAAAT
CCAAGCGAAAAGTGGCATTCTATCAAGTGACATTGAATGTGGTCTCATATTAATAACTGT
NNNAGCAAGAACTGGAAGTGAAAGAAGAAGAAGAACAGCTGTAATAACAACAGC


backtranseqambig with invertebrate mitochondrial:

>EMBOSS_001
YTNYTNGGNGAYGAYCARYTNTAYAAYGTNATYGTNACNGCNCAYGCNTTYGTNATRATY
TTYTTYATRGTNATRCCNATYYTNATYGGNGGNTTYGGNAAYTGRYTNGTNCCNYTNATR
YTNGGNGCNCCNGAYATRGCNTTYCCNCGNYTNAAYAAYYTNWSNTTYTGRATRYTNCCN
CCNWSNYTNACNYTNYTNGTNGCNWSNWSNATRGTNGARWSNGGNGTNGGNACNGGNTGR
ACNGTNTAYCCNCCNYTNWSNGCNGCNATYGCNCAYGCNGGNCCNWSNGTNGAYYTNGCN
ATYTTYWSNYTNCAYYTNGCNGGNATYWSNWSNATYYTNGGNGCNGTNAAYTTYATYACN
ACNGTNATYAAYATRCGNCCNCAYWSNATRWSNYTNGAYCGNATRCCNYTNTTYGCNTGR
GCNGTNGTNATYACNGCNGTNYTNYTNYTNYTNWSNYTNCCNGTNYTNGCNNNN

```{r}

```

# 3.1

C. elegans was the first multicellular organism to have its whole genome sequenced, and in 2019[14] it was the first organism to have its connectome (neuronal "wiring diagram") completed.