---
title: "LAB 1 Bioinformatics"
date: "`r Sys.Date()`"
author: 
  - Hugo Morvan
  - William Wiik
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
  html_document:
    df_print: paged
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A96 Advanced Machine Learning \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\
  Department of Computer and Information Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r, include = FALSE,message=FALSE}

# Q1 ###########################################################################
set.seed(12345)

library(ggplot2)
library(dplyr)
```


# Question 1

We consider a gene locus with two possible alleles (say $A$ and $a$) and a diploid population with
$N$ individuals. Hence, there are 2$N$ alleles in the population. Let $p$ be the proportion of $A$s in
the allele population and $q$ the population of as (of course $p$ + $q$ = 1). A population is said to
be in Hardy-Weinberg equilibrium if the proportion of $AA$ homozygotes is $p^2$, aa homozygotes
is $q^2$ and the proportion of heterozygotes ($Aa$) is $2pq$.


<!-- alleles = gener -->
<!-- mating = parning -->


## Question 1.1

Show that with random mating (i.e. both alleles of the o spring are just randomly, with proportions $p$ and $q$, drawn from the parental allele population) Hardy-Weinberg equilibrium is attained
in the first generation. What is the proportion of $A$ and $a$ alleles in the offspring population?
Hence, with random mating, can a population in Hardy-Weinberg equilibrium ever deviate from
it?

If the mating is random then all individuals inherits alleles indepnedently with probabilities $p$ and $q$ from
each parent. Then:

$$P(AA) = p\cdot p =p^2$$
$$P(aa) = q\cdot q= q^2$$
$$P(Aa) = (p \cdot q) + (p \cdot q) = 2pq$$

<!-- ?? -->

The population can not deviate from Hardy-Weinberg equilibrium unless factors such as mutation changes the alleles.

<!-- ?? -->

```{r}


```



## Question 1.2

We look at the MN blood group, it
has two possible codominating (both contribute to heterozygotes) alleles $L^M$ (denoted $M$) and
$L^N$ (denoted $N$). In a population of 1000 Americans of Caucasian descent the following genotype
counts were observed, 357 individuals were $MM$, 485 were $MN$ and 158 were $NN$. Use a chisquare goodness of t test to test if the population is in Hardy-Weinberg equilibrium.


```{r}

MM <- 357
MN <- 485
NN <- 158

population <- 1000

# p = the proportion of Ms in the allele population 
p <- (2 * MM + MN) / (2 * population)
q <- 1 - p

# Hardy-Weinberg equilibrium
HW_MM <- p**2  * population
HW_NN <- q**2  * population
HW_MN <- q * p  * population * 2


# chisq = sum[(Observed - Expected)^2 / (Expected)]
##pchisq()


```


\newpage


# Question 2

In this exercise, you will use ...

```{r}


```




