---
title: "Escherichia Coli Network"
output: pdf_document
author: ""
date: Example for GeneNet 1.2.13 (August 2015) or later
---

This note reproduces the “Escherichia coli” network example from J. Schäfer and 
K. Strimmer. 2005. *A shrinkage approach to large-scale covariance 
estimation and implications for functional genomics.* 
Statist. Appl. Genet. Mol. Biol. **4**: 32 
(http://dx.doi.org/10.2202/1544-6115.1175)


# Load GeneNet package
```{r}
library("GeneNet")
```



E. Coli data set (9 time points for 102 genes):

```{r}
data(ecoli)
dim(ecoli)
```


# Estimation of partial correlations

Estimate matrix of partial correlation using a shrinkage estimator:

```{r}
pc = ggm.estimate.pcor(ecoli)
dim(pc)
```

Assign p-values, q-values and empirical posterior probabilities to all
5151 potential edges in the network:

```{r}
ecoli.edges = network.test.edges(pc, direct=TRUE, fdr=TRUE)
dim(ecoli.edges)
```

The table lists all edges in the order strength of partial correlations:

```{r}
ecoli.edges[1:5,]
```

# Decide which edges to include in the network

To obtain a graph you need to select top ranking edges according to a suitable criterion.  Here are some suggestions:

1. Use local fdr cutoff 0.2, i.e. include all edges with posterior probability of at least 0.8.

```{r}
ecoli.net = extract.network(ecoli.edges)
dim(ecoli.net)
```

2. Use local fdr cutoff 0.1, i.e. i.e. include all edges with posterior probability of at least 0.9.


```{r}
ecoli.net = extract.network(ecoli.edges, cutoff.ggm=0.9, cutoff.dir=0.9)
dim(ecoli.net)
```

3. Include a fixed number of edges, say the 70 strongest edges

```{r}
ecoli.net = extract.network(ecoli.edges, method.ggm="number", cutoff.ggm=70)
dim(ecoli.net)
```


Plot network

For plotting we use the graph and Rgraphviz packages from Bioconductor.

```{r}
library("Rgraphviz") 
```

Create graph object from the list of edges:

```{r}
node.labels = colnames(ecoli)
gr = network.make.graph(ecoli.net, node.labels, drop.singles=TRUE)
table(  edge.info(gr)$dir )
sort( node.degree(gr), decreasing=TRUE)
```


Set node and edge attributes for more beautiful graph plotting:

```{r}
globalAttrs = list()
globalAttrs$edge = list(color = "black", lty = "solid", lwd = 1, arrowsize=1)
globalAttrs$node = list(fillcolor = "lightblue", shape = "ellipse", fixedsize = FALSE)
 
nodeAttrs = list()
nodeAttrs$fillcolor = c('sucA' = "yellow")

edi = edge.info(gr)
edgeAttrs = list()
edgeAttrs$dir = edi$dir # set edge directions 
edgeAttrs$lty = ifelse(edi$weight < 0, "dotted", "solid") # negative correlation -> dotted
edgeAttrs$color = ifelse(edi$dir == "none", "black", "red")
edgeAttrs$label = round(edi$weight, 2) # use partial correlation as edge labels
```


```{r fig.width=8, fig.height=7}
plot(gr, attrs = globalAttrs, nodeAttrs = nodeAttrs, edgeAttrs = edgeAttrs, "fdp")

```

