---
title: "LAB 5 Bioinformatics"
author:
- Hugo Morvan
- William Wiik
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    number_sections: true
  html_document:
    df_print: paged
  word_document: default
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A51 Bioinformatics \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\ Department of Computer and Information
  Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r, include = FALSE,message=FALSE}
set.seed(12345)

```


# Question 1

Go to the webpage http://snap.stanford.edu/biodata/ and choose one of the provided datasets. Download it and reproduce the statistics concerning the graph. If you obtain different values, then discuss this in your report. Visualize the graph. The next step is to try to identify some clusters (communities in the graph). You can follow the tutorial at https://psych-networks.com/r-tutorial-identify-communities-items-networks/ to achieve this. Once you have found some clusters, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts. If it will not be possible to do this question on the whole downloaded graph, then you may take some sub-graph of it.


```{r,message=FALSE}

library(qgraph)
library(igraph)
library(ape)


con <- gzfile("X:/Desktop/Bioinformatics/Lab5/ChG-Miner_miner-chem-gene.tsv.gz", "rt")
data <- read.table(con, sep = "\t", header = FALSE)
close(con)
colnames(data) <- c("Drug","Gene")

g <- graph_from_data_frame(data, directed = TRUE)
g


# Number of nodes and edges
cat("Number of nodes:", vcount(g), "\n")
cat("Number of edges:", ecount(g), "\n")


```


```{r,message=FALSE}


plot(g, vertex.size=5,vertex.label=NA, edge.width=0.01,vertex.color="coral", main="ChG-Miner_miner-chem-gene Graph")

```

```{r plot1, fig.width=16, fig.height=16, cache=TRUE}

# Function to create and plot a bipartite network from an edge matrix
plot_bipartite_network <- function(edge_matrix, title = "Bipartite Network",plot_graph = TRUE, top_n = 9) {
  # Create unique node names
  drugs <- unique(edge_matrix[,1])
  genes <- unique(edge_matrix[,2])
  
  # Create the graph
  g <- make_empty_graph(directed = FALSE)
  
  # Add all nodes
  all_nodes <- c(drugs, genes)
  g <- add_vertices(g, length(all_nodes), name = all_nodes)
  
  # Create type attribute for bipartite graph
  V(g)$type <- c(rep(TRUE, length(drugs)), rep(FALSE, length(genes)))
  

  edges <- apply(edge_matrix, 1, function(x) c(which(V(g)$name == x[1]), which(V(g)$name == x[2])))
  
  # Add all edges at once
  g <- add_edges(g, unlist(edges))

  # Set node colors based on type
  #V(g)$color <- ifelse(V(g)$type, "lightblue", "lightgreen")
  
  comms <- cluster_louvain(g)
  freq_memb <- table(membership(comms))
  big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:top_n]

  right_index <- as.vector(as.numeric(names(big_mems)))
  mems <- membership(comms)
  comm_col <- rainbow(top_n)[mems]
  comm_col <- rep("#FFFFFF",length(comm_col))

  topncol <- c("#FF0000", "#FFAA00", "#AAFF00", "#00FF00" ,"#00FFAA",
               "#00AAFF", "#0000FF" ,"#AA00FF", "#FF00AA")

  for(i in 1:length(right_index)) {
    comm_col[which(mems == right_index[i])] <- topncol[i]
  }
  


  V(g)$shape <- ifelse(V(g)$type, "circle", "square")
  # Plot the network
  if(plot_graph) {
    
    plot(g, 
         vertex.label = V(g)$name,
         vertex.shape = V(g)$shape,
         vertex.color = comm_col,
         vertex.label.cex = 0.01,  # Adjust label size
         vertex.size = 5,         # Node size
         vertex.label.color = "black",
         main = title#,
         #layout = layout_as_bipartite  # Specific layout for bipartite graphs
    )
    legend("bottomright", legend = paste("Community", right_index),
           fill = topncol)

    # Return the graph object for further analysis if needed
    return(g)
  } else {
    return(comms)
  }
}

# Example usage:
# Create a sample edge matrix (drug-gene connections)
edge_matrix <- data

# Plot the network
plot_bipartite_network(edge_matrix, "Drug-Gene Interaction Network")
```

## Graph 2


```{r plot2, fig.width=16, fig.height=16, cache=TRUE}


# Function to create and plot a bipartite network from an edge matrix
plot_bipartite_network_2 <- function(edge_matrix, title = "Bipartite Network", top_n = 9) {
  # Create unique node names
  drugs <- unique(edge_matrix[,1])
  genes <- unique(edge_matrix[,2])
  
  # Create the graph
  g <- make_empty_graph(directed = FALSE)
  
  # Add all nodes
  all_nodes <- c(drugs, genes)
  g <- add_vertices(g, length(all_nodes), name = all_nodes)
  
  # Create type attribute for bipartite graph
  V(g)$type <- c(rep(TRUE, length(drugs)), rep(FALSE, length(genes)))
  
  # Add edges
  edges <- apply(edge_matrix, 1, function(x) c(which(V(g)$name == x[1]), which(V(g)$name == x[2])))
  
  # Add all edges at once
  g <- add_edges(g, unlist(edges))
  # Set node colors based on type
  #V(g)$color <- ifelse(V(g)$type, "lightblue", "lightgreen")
  
  comms <- cluster_louvain(g)
  freq_memb <- table(membership(comms))
  big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:top_n]

  right_index <- as.vector(as.numeric(names(big_mems)))
  mems <- membership(comms)
  comm_col <- rainbow(top_n)[mems]
  comm_col <- rep("#FFFFFF",length(comm_col))

  topncol <- c("#FF0000", "#FFAA00", "#AAFF00", "#00FF00" ,"#00FFAA",
               "#00AAFF", "#0000FF" ,"#AA00FF", "#FF00AA")

  for(i in 1:length(right_index)) {
    comm_col[which(mems == right_index[i])] <- topncol[i]
  }
  

  # Plot the network
  
  plot(g, 
       vertex.label = V(g)$name,
       vertex.color = comm_col,
       vertex.label.cex = 0.1,  # Adjust label size
       vertex.size = 5,         # Node size
       vertex.label.color = "black",
       main = title,
       layout = layout_as_bipartite  # Specific layout for bipartite graphs
  )
  legend("bottomright", legend = paste("Community", right_index),
           fill = topncol)
  # Return the graph object for further analysis if needed
  return(g)
}

# Example usage:
# Create a sample edge matrix (drug-gene connections)
edge_matrix <- data

# Plot the network
plot_bipartite_network_2(edge_matrix, "Drug-Gene Interaction Network")
```



```{r first_com}
top_n <- 9
communitys <- plot_bipartite_network(edge_matrix, plot_graph = FALSE)
freq_memb <- table(membership(communitys))
big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:top_n]

right_index <- as.vector(as.numeric(names(big_mems)))



biggest_com <- communitys[[right_index[1]]]


# install.packages("BiocManager")
# BiocManager::install("clusterProfiler")
# BiocManager::install("org.Hs.eg.db")
# devtools::install_github("YuLab-SMU/clusterProfiler")

library(clusterProfiler)
library(org.Hs.eg.db)

# Perform GO enrichment analysis for the current community
go_results <- enrichGO(gene = biggest_com,
                       OrgDb = org.Hs.eg.db, keyType = "UNIPROT", ont = "BP")

top_go_terms <- go_results@result[order(go_results@result$p.adjust), ]
head(top_go_terms$Description, 10) # strongly related to neurobiology
barplot(go_results, showCategory = 10)



```



```{r sec_com}

sec_biggest_com <- communitys[[right_index[2]]]


# Perform GO enrichment analysis for the current community
go_results <- enrichGO(gene = sec_biggest_com,
                       OrgDb = org.Hs.eg.db, keyType = "UNIPROT", ont = "BP")

top_go_terms <- go_results@result[order(go_results@result$p.adjust), ]
head(top_go_terms$Description, 10) 
barplot(go_results, showCategory = 10)


```


# Question 2


Recreate one of the three analyses that can be found on https://strimmerlab.github.io/software/genenet/index.html. Document and discuss all your steps. In the analyses there is the step where you select the edges to keep. There a particular criterion is chosen for edge inclusion. Vary this criterion and explore how the resulting clusters will differ with the changes. Take one found cluster, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts.







