---
title: "LAB 5 Bioinformatics"
author:
- Hugo Morvan
- William Wiik
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    number_sections: true
  html_document:
    df_print: paged
  word_document: default
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A51 Bioinformatics \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\ Department of Computer and Information
  Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r, include = FALSE,message=FALSE}
set.seed(12345)

```


# Question 1

Go to the webpage http://snap.stanford.edu/biodata/ and choose one of the provided datasets. Download it and reproduce the statistics concerning the graph. If you obtain different values, then discuss this in your report. Visualize the graph. The next step is to try to identify some clusters (communities in the graph). You can follow the tutorial at https://psych-networks.com/r-tutorial-identify-communities-items-networks/ to achieve this. Once you have found some clusters, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts. If it will not be possible to do this question on the whole downloaded graph, then you may take some sub-graph of it.


```{r,message=FALSE}

library(qgraph)
library(igraph)
library(ape)

# Test 1

#https://snap.stanford.edu/biodata/datasets/10001/10001-ChCh-Miner.html
con <- gzfile("X:/Desktop/Bioinformatics/Lab5/ChCh-Miner_durgbank-chem-chem.tsv.gz", "rt")
data <- read.table(con, sep = "\t", header = FALSE)
colnames(data) <- c("Drug1","Drug2")
close(con)


# Test 2

# con <- gzfile("X:/Desktop/Bioinformatics/Lab5/SS-Butterfly_weights.tsv.gz", "rt")
# data <- read.table(con, sep = "\t", header = TRUE)
# close(con)
# graph_data <- data.frame(from = data$X0, to = data$X4, weight = data$X0.029663)
# head(data)
# 
# con <- gzfile("X:/Desktop/Bioinformatics/Lab5/SS-Butterfly_labels.tsv.gz", "rt")
# data_labels <- read.table(con, sep = "\t", header = TRUE)
# close(con)
# head(data_labels)


# Test 3

con <- gzfile("X:/Desktop/Bioinformatics/Lab5/ChG-Miner_miner-chem-gene.tsv.gz", "rt")
data <- read.table(con, sep = "\t", header = FALSE)
close(con)
colnames(data) <- c("Drug","Gene")

g <- graph_from_data_frame(data, directed = TRUE)
g


# Number of nodes and edges
cat("Number of nodes:", vcount(g), "\n")
cat("Number of edges:", ecount(g), "\n")


```


```{r,message=FALSE, fig.w }

# TEST 1
# plot(g, vertex.size=5, vertex.label=NA, edge.width=0.01,vertex.color="coral", main="ChCh-Miner Graph")

# TEST 2
# V(g)$label <- as.character(data_labels$X1[match(V(g)$name, data_labels$X0)])
# plot(g, vertex.size = 10, vertex.label.cex = 0.7,
#      edge.width = E(g)$weight,
#      vertex.label = V(g)$label,
#      main = "Butterfly Similarity Network")


# TEST 3
plot(g, vertex.size=5,vertex.label=NA, edge.width=0.01,vertex.color="coral", main="ChG-Miner_miner-chem-gene Graph")

```



```{r,message=FALSE}
#  TEST 1

# library(qgraph)
# 
# edges <- data.frame(Source = data$DB00862, Target = data$DB00966)
# 
# # List of unique drugs
# drug_ids <- unique(c(as.character(data$DB00862), as.character(data$DB00966)))
# 
# # Initialize an empty adjacency matrix
# adj_matrix <- matrix(0, nrow = length(drug_ids), ncol = length(drug_ids))
# rownames(adj_matrix) <- colnames(adj_matrix) <- drug_ids
# 
# # Fill the adjacency matrix based on the drug pairs in the edges list
# for (i in 1:nrow(edges)) {
#   adj_matrix[edges$Source[i], edges$Target[i]] <- 1
#   adj_matrix[edges$Target[i], edges$Source[i]] <- 1  # Since it's undirected
# }
# cormatrix <- cor(adj_matrix)
# 
# 
# graph1<-qgraph(cormatrix, layout="spring", sampleSize = nrow(data),
#               vsize=2, cut=0, maximum=.45, border.width=1.5)
# graph1


```

```{r,message=FALSE}
# TEST 2
# butterfly_ids <- unique(c(as.character(data$X0), as.character(data$X4)))
# 
# # Create an empty adjacency matrix initialized to 0
# adj_matrix <- matrix(0, nrow = length(butterfly_ids), ncol = length(butterfly_ids))
# rownames(adj_matrix) <- colnames(adj_matrix) <- butterfly_ids
# 
# # Fill the adjacency matrix with similarity scores
# for (i in 1:nrow(data)) {
#   adj_matrix[as.character(data$X0[i]), as.character(data$X4[i])] <- data$X0.029663[i]
#   adj_matrix[as.character(data$X4[i]), as.character(data$X0[i])] <- data$X0.029663[i]  # Symmetric matrix
# }
# 
# cor_matrix <- cor(adj_matrix)
# qgraph(cor_matrix, layout="spring", sampleSize = nrow(data),
#               vsize=7, cut=0, maximum=.45, border.width=1.5)

```

```{r,message=FALSE}
# TEST 3
# butterfly_ids <- unique(c(as.character(data$X0), as.character(data$X4)))
# 
# # Create an empty adjacency matrix initialized to 0
# adj_matrix <- matrix(0, nrow = length(butterfly_ids), ncol = length(butterfly_ids))
# rownames(adj_matrix) <- colnames(adj_matrix) <- butterfly_ids
# 
# # Fill the adjacency matrix with similarity scores
# for (i in 1:nrow(data)) {
#   adj_matrix[as.character(data$X0[i]), as.character(data$X4[i])] <- data$X0.029663[i]
#   adj_matrix[as.character(data$X4[i]), as.character(data$X0[i])] <- data$X0.029663[i]  # Symmetric matrix
# }
# 
# cor_matrix <- cor(adj_matrix)
# qgraph(cor_matrix, layout="spring", sampleSize = nrow(data),
#               vsize=7, cut=0, maximum=.45, border.width=1.5)

```


```{r,message=FALSE}


# Community detection using Walktrap method
community_walktrap <- cluster_walktrap(g)

# Visualize the communities (coloring nodes by their community)
plot(community_walktrap, g, vertex.size=5, vertex.label=NA, main="Walktrap Community Detection")


devtools::install_github('hfgolino/EGA')
library(EGAnet)


data_mini <- data[1:1000,]

# TEST 3
unique_nodes <- unique(c(as.character(data_mini$Drug), as.character(data_mini$Gene)))

# Create an empty adjacency matrix initialized to 0
adj_matrix <- matrix(0, nrow = length(ids), ncol = length(ids))
rownames(adj_matrix) <- colnames(adj_matrix) <- ids

# Fill the adjacency matrix with similarity scores
for (i in 1:nrow(data_mini)) {
  adj_matrix[as.character(data_mini$Drug[i]), as.character(data_mini$Gene[i])] <- 1
  adj_matrix[as.character(data_mini$Gene[i]), as.character(data_mini$Drug[i])] <- 1
}

cov(data_mini)
cormatrix <- cor_auto(data)
graph1<-qgraph(adj_matrix, layout="circular", sampleSize = nrow(data_mini),
              vsize=7, cut=0, maximum=.45, border.width=1.5)


g2 <- as.undirected(g, mode = "collapse")
community <- cluster_louvain(g2)
comm_membership <- membership(community)
community_colors <- rainbow(length(unique(comm_membership)))[comm_membership]
plot(g, 
     vertex.size = 5,                    # Adjust vertex size for better visibility
     vertex.color = community_colors,     # Color nodes by their community
     vertex.label = NA,                   # Hide labels for clarity (optional)
     edge.arrow.size = 0.5,               # Adjust edge size
     main = "Community Detection Plot")

```

# Question 2


Recreate one of the three analyses that can be found on https://strimmerlab.github.io/software/genenet/index.html. Document and discuss all your steps. In the analyses there is the step where you select the edges to keep. There a particular criterion is chosen for edge inclusion. Vary this criterion and explore how the resulting clusters will differ with the changes. Take one found cluster, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts.







