---
title: "LAB 5 Bioinformatics"
author:
- Hugo Morvan
- William Wiik
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    number_sections: true
  html_document:
    df_print: paged
  word_document: default
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A51 Bioinformatics \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\ Department of Computer and Information
  Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
editor_options: 
  chunk_output_type: console
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r, include = FALSE,message=FALSE}
set.seed(12345)

```

# Question 2


Recreate one of the three analyses that can be found on https://strimmerlab.github.io/software/genenet/index.html. Document and discuss all your steps. In the analyses there is the step where you select the edges to keep. There a particular criterion is chosen for edge inclusion. Vary this criterion and explore how the resulting clusters will differ with the changes. Take one found cluster, identify the elements in it and try to find information on this cluster. Is it related to some known biological phenomena? If you do not find anything, then document your search attempts.
```{r data}
con <- gzfile("Lab5/ChG-Miner_miner-chem-gene.tsv.gz", "rt")
#con <- gzfile("ChG-Miner_miner-chem-gene.tsv.gz", "rt")
data <- read.table(con, sep = "\t", header = FALSE)
close(con)
colnames(data) <- c("Drug","Gene")
```

## Graph 1

```{r plot1, fig.width=16, fig.height=16, cache=TRUE}
library(igraph)

# Function to create and plot a bipartite network from an edge matrix
plot_bipartite_network <- function(edge_matrix, title = "Bipartite Network",plot_graph = TRUE, top_n = 9) {
  # Create unique node names
  drugs <- unique(edge_matrix[,1])
  genes <- unique(edge_matrix[,2])
  
  # Create the graph
  g <- make_empty_graph(directed = FALSE)
  
  # Add all nodes
  all_nodes <- c(drugs, genes)
  g <- add_vertices(g, length(all_nodes), name = all_nodes)
  
  # Create type attribute for bipartite graph
  V(g)$type <- c(rep(TRUE, length(drugs)), rep(FALSE, length(genes)))
  
  # Add edges
  for(i in 1:nrow(edge_matrix)) {
    g <- add_edges(g, 
                   c(which(V(g)$name == edge_matrix[i,1]), 
                     which(V(g)$name == edge_matrix[i,2])))
  }
  
  # Set node colors based on type
  #V(g)$color <- ifelse(V(g)$type, "lightblue", "lightgreen")
  
  comms <- cluster_louvain(g)
  freq_memb <- table(membership(communitys))
  big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:top_n]

  right_index <- as.vector(as.numeric(names(big_mems)))
  mems <- membership(comms)
  comm_col <- rainbow(top_n)[mems]
  comm_col <- rep("#FFFFFF",length(comm_col))

  topncol <- c("#FF0000", "#FFAA00", "#AAFF00", "#00FF00" ,"#00FFAA",
               "#00AAFF", "#0000FF" ,"#AA00FF", "#FF00AA")

  for(i in 1:length(right_index)) {
    comm_col[which(mems == right_index[i])] <- topncol[i]
  }
  


  V(g)$shape <- ifelse(V(g)$type, "circle", "square")
  # Plot the network
  if(plot_graph) {
    plot(g, 
         vertex.label = V(g)$name,
         vertex.shape = V(g)$shape,
         vertex.color = comm_col,
         vertex.label.cex = 0.01,  # Adjust label size
         vertex.size = 5,         # Node size
         vertex.label.color = "black",
         main = title#,
         #layout = layout_as_bipartite  # Specific layout for bipartite graphs
    )
    legend("bottomright", legend = paste("Community", right_index),
           fill = topncol)
    # Return the graph object for further analysis if needed
    return(g)
  } else {
    return(comms)
  }
}

# Example usage:
# Create a sample edge matrix (drug-gene connections)
edge_matrix <- data

# Plot the network
plot_bipartite_network(edge_matrix, "Drug-Gene Interaction Network")
```

## Graph 2


```{r plot2, fig.width=16, fig.height=16, cache=TRUE}
library(igraph)

# Function to create and plot a bipartite network from an edge matrix
plot_bipartite_network <- function(edge_matrix, title = "Bipartite Network", top_n = 9) {
  # Create unique node names
  drugs <- unique(edge_matrix[,1])
  genes <- unique(edge_matrix[,2])
  
  # Create the graph
  g <- make_empty_graph(directed = FALSE)
  
  # Add all nodes
  all_nodes <- c(drugs, genes)
  g <- add_vertices(g, length(all_nodes), name = all_nodes)
  
  # Create type attribute for bipartite graph
  V(g)$type <- c(rep(TRUE, length(drugs)), rep(FALSE, length(genes)))
  
  # Add edges
  for(i in 1:nrow(edge_matrix)) {
    g <- add_edges(g, 
                   c(which(V(g)$name == edge_matrix[i,1]), 
                     which(V(g)$name == edge_matrix[i,2])))
  }
  
  # Set node colors based on type
  #V(g)$color <- ifelse(V(g)$type, "lightblue", "lightgreen")
  
  comms <- cluster_louvain(g)
  freq_memb <- table(membership(communitys))
  big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:top_n]

  right_index <- as.vector(as.numeric(names(big_mems)))
  mems <- membership(comms)
  comm_col <- rainbow(top_n)[mems]
  comm_col <- rep("#FFFFFF",length(comm_col))

  topncol <- c("#FF0000", "#FFAA00", "#AAFF00", "#00FF00" ,"#00FFAA",
               "#00AAFF", "#0000FF" ,"#AA00FF", "#FF00AA")

  for(i in 1:length(right_index)) {
    comm_col[which(mems == right_index[i])] <- topncol[i]
  }
  

  # Plot the network
  
  plot(g, 
       vertex.label = V(g)$name,
       vertex.color = comm_col,
       vertex.label.cex = 0.1,  # Adjust label size
       vertex.size = 5,         # Node size
       vertex.label.color = "black",
       main = title,
       layout = layout_as_bipartite  # Specific layout for bipartite graphs
  )
  legend("bottomright", legend = paste("Community", right_index),
           fill = topncol)
  # Return the graph object for further analysis if needed
  return(g)
}

# Example usage:
# Create a sample edge matrix (drug-gene connections)
edge_matrix <- data

# Plot the network
plot_bipartite_network(edge_matrix, "Drug-Gene Interaction Network")
```


```{r poop}

communitys <- plot_bipartite_network(edge_matrix, plot_graph = FALSE)
length(unique(communitys))
freq_memb <- table(membership(communitys))
big_mems <- freq_memb[order(freq_memb,decreasing = T)][1:9]

gg <- communitys[as.vector(as.numeric(names(big_mems)))]
membgg <- membership(gg)



memb <- membership(communitys)
go_results_list <- list()

# Loop through each community and perform GO enrichment
for (i in 1:length(communitys$membership)) {
  # Get the genes in the current community
  community_genes <- communitys$names[communitys$membership == i]
  
  # Perform GO enrichment analysis for the current community
  go_results <- enrichGO(
    gene = community_genes, 
    OrgDb = org.Hs.eg.db, 
    keyType = "UNIPROT",  # Use "UNIPROT" or the appropriate gene ID type
    ont = "BP"  # Change "BP" to "MF" or "CC" if you want other ontologies
  )
  
  # Store the results in the list
  go_results_list[[i]] <- go_results
}

```
