---
title: "LAB 3 Bioinformatics"
author:
- Hugo Morvan
- William Wiik
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    fig_caption: true
    number_sections: true
  html_document:
    df_print: paged
  word_document: default
geometry: top=100pt,bottom=100pt,left=68pt,right=66pt
header-includes:
- \usepackage{float}
- \usepackage{longtable}
- \usepackage{caption}
- \usepackage{fancyhdr}
- \usepackage{titling}
- \renewcommand{\headrulewidth}{0pt}
- \renewcommand{\and}{\\}
- \pretitle{\centering\vspace{0cm}{732A51 Bioinformatics \par}\vspace{5cm}\Huge\textbf}
- \posttitle{\vspace{1cm}\large\textbf{}\par}
- \preauthor{\centering\vspace{4cm}\normalsize}
- \postauthor{\par\vspace{2cm}}
- \predate{\centering{\normalsize STIMA \\ Department of Computer and Information
  Science \\ Linköpings universitet \par}}
- \postdate{\par\vspace{0cm}}
- \raggedbottom
---


<!-- page number pos -->
\fancyhf{}
\fancyfoot[C]{\thepage}
\pagestyle{fancy}

<!-- no page nr on first page  -->
\pagenumbering{gobble}

<!-- Anger sidbrytning -->
\clearpage

<!-- creating the table of contents -->
\setcounter{tocdepth}{3}
\tableofcontents

<!-- new page -->
\clearpage

<!-- starting the count on 1 after the contents -->
\pagenumbering{arabic}
\setcounter{page}{1}

```{r, include = FALSE,message=FALSE}
set.seed(12345)
library(dplyr)

```


# Question 1

Using the script http://ape-package.ird.fr/APER/APER2/SylviaWarblers.R obtain the *Sylvia*
warblers phylogeny (the script saves in in the file sylvia nj k80.tre). The geographical range
data can be found in http://ape-package.ird.fr/APER/APER2/sylvia_data.txt and in the
script is referenced as `DF$geo.range`. Notice that one tip is removed due to missing data


```{r 1-0}
# 
# tr <- drop.tip(tr, "Chamaea_fasciata")
# DF <- sylvia.eco[tr$tip.label, ]


```


## Question 1.1

Explain all the steps in the script required to obtain the phylogeny and trait data.


**Answer:**

```{r 1-1}

```


## Question 1.2 *

```{r 1-2}

```


\newpage


# Question 2

Install the `ade4` package. Included with it you will find the carnivores dataset, `data(carni70)`

## Question 2.1

Explore the data set and report what can be found in it. Provide some plots.


```{r 2-1}
library(ade4)
library(ggplot2)
library(cowplot)

data(carni70)

tab_df <- as.data.frame(carni70$tab)
summary(tab_df)


p1 <- ggplot(tab_df, aes(x = size)) + 
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Size", x = "Size", y = "Frequency") + theme_bw()

p2 <- ggplot(tab_df, aes(x = range)) + 
  geom_histogram(binwidth = 1, fill = "salmon", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Range", x = "Range", y = "Frequency") + theme_bw()

p3 <- ggplot(tab_df, aes(x = size, y = range)) + 
  geom_point(color = "blue") + 
  labs(title = "Scatter Plot of Size vs Range", x = "Size", y = "Range") + theme_bw()


cowplot::plot_grid(p1, p2, p3, ncol = 2)  




```

```{r, echo=F}

print(paste("Total number of carnivores:",nrow(tab_df)))

print(paste("The carnivore with biggest size:",rownames(tab_df)[which.max(tab_df$size)]))
print(paste("The carnivore with smallest size:",rownames(tab_df)[which.min(tab_df$size)]))

print(paste("The carnivore with biggest range:",rownames(tab_df)[which.max(tab_df$range)]))
print(paste("The carnivore with smallest range:",rownames(tab_df)[which.min(tab_df$range)]))
```
There are 70 carnivores, with a median size of 3.2 and a median range of 6.1. Two clear outliers in size are Ursus arctos (Brown bear) and Tremarctos ornatus (Spectacled bear).



## Question 2.2*
